---
date: "2021-09-21"
title: Lake Monona & Exoplanets (R)
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE)
library(tidyverse)
library(lubridate)
library(scales)
```
    
```{r read-data, echo = FALSE}
## The echo = FALSE argument to this chunk
##   means that the code will not appear in the output
##   but the code will run so that the data is in the session

## Read Lake Monona data
## Remove rows with missing data
monona = read_csv("lake-monona-winters-2021.csv") %>%
  drop_na()
```

In the following code I read in a csv file that contained data about Lake Monona in Madison, Wisconsin and created a histogram to display winter durations. 

```{r}
ggplot(monona) +
  geom_histogram(aes(x = duration), 
  fill = "darkslategray1",
  color = "black",
  binwidth = 5,
  center = 70) +
  xlab("Days Frozen") +
  ylab("Occurances (Years)") +
  ggtitle("Lake Monona: Freeze Durations: 1855-2020")
```

Next, I created a line plot and estimated the winter duration trend through the years.

```{r}
ggplot(monona, aes(x = year1, y = duration)) +
  geom_line() +
  geom_smooth(se = FALSE) +
  geom_smooth(method="lm", se=FALSE, color = "red") +
  xlab("Year") +
  ylab("Days Frozen") +
  ggtitle("Lake Monona: Freeze Durations: 1855-2020")
```

```{r echo = FALSE}
## Fit the model
fit = lm(duration ~ year1, data = monona)

## Add fitted values and residuals to the data set
monona = monona %>% 
  mutate(fitted = fitted(fit),
         residuals = residuals(fit))
```

Then, I created a variable for the residials of the freeze durations. The code below creates a plot of said residuals versus each year with another trend line.

```{r}
ggplot(monona, aes(x = year1, y = residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_smooth(se = FALSE) +
  xlab("Year") +
  ylab("Residuals: Days Frozen") + 
  ggtitle("Lake Monona: Freeze Durations: 1855-2020")
```

I made a new variable again; this time to count the number of days after June 30 on which Lake Monona first freezes. Below is a histogram that displays this new data:

```{r echo = FALSE}
## add winter_day = #days of first freeze after June 30
monona = monona %>% 
  mutate(winter_day = as.numeric(
    first_freeze - ymd(str_c(year1,"-06-30"))))
```

```{r}
ggplot(monona, aes(x = winter_day)) +
  geom_histogram(binwidth = 5, center = 154, color = "black", fill = "bisque") +
  scale_x_continuous(
    breaks = c(154, 169, 184, 199, 214),
    label = c("Dec 1", "Dec 16", "Dec 31", "Jan 15", "Jan 30")) +
  xlab("Days after June 30") +
  ylab("Count") +
  ggtitle("Lake Monona: First Freeze Distribution",
          subtitle = "1855 - 2021")
```

I switched gears and read in data from a csv file about exoplanets. Below I modified the data and created a couple graphs.

```{r}
exo = read_csv("exoplanets_default_2021-06-15.csv", skip = 100) %>% 
  select(pl_name, hostname, discoverymethod, pl_bmasse) %>%
  rename(planet = pl_name,
         star = hostname,
         method = discoverymethod,
         mass = pl_bmasse) %>% 
  drop_na(mass)
```

```{r}
ggplot(exo, aes(x = method, y = mass)) +
  geom_boxplot() +
  scale_y_continuous(trans = "log10", label = comma) +
  xlab("Method of Detection") +
  ylab("Mass (Earth masses)") +
  coord_flip()
```


```{r}
ggplot(exo) +
  geom_density(aes(x = mass),
             fill = "blue",
             color = "black") +
  xlab("Earth Masses")  +
  ylab("Density") +
  ggtitle("Planet Mass Density") +
scale_x_continuous(trans = "log10", label = comma)  
```


