---
date: "2021-10-10"
title: Airports (R)
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, fig.height = 3)
library(tidyverse)
library(scales)
library(lubridate)
source("viridis.R")
```

In the following code, I read in csv files that contained data from the Chicago O'Hare airport from 2016-2020. Then, I:
- Combined the files into one data set
- Added columns for the year, month, day, and day of the week
- Renamed/removed a few variables
- Rearanged the columns
- Printed the first 10 rows

```{r}
ord16 <- read_csv("ORD-2016.csv")
ord17 <-  read_csv("ORD-2017.csv")
ord18 <-  read_csv("ORD-2018.csv")
ord19 <-  read_csv("ORD-2019.csv")
ord20 <-  read_csv("ORD-2020.csv")
ord <- bind_rows(ord16, ord17, ord18, ord19, ord20)

ord1 <- ord %>% 
  select(-terminal) %>%
  rename(passengers = all_total,
         flights = all_flights,
         booths = all_booths) %>%
  mutate(year = year(date)) %>% 
  mutate(month = month(date, label = TRUE, abbr = TRUE)) %>% 
  mutate(day = day(date)) %>% 
  mutate(wday = wday(date, label = TRUE, abbr = FALSE)) %>% 
select(date, year, month, day, wday, everything()) %>% 
  arrange(date, hour)
  
(head(ord1, n = 10))

rm(ord16, ord17, ord18, ord19, ord20)
```

Next I checked if there were any missing dates in the range from January 1, 2016 to December 31, 2020 that are missing? Sure enough, there were eight of them:

```{r}
possible_dates <- seq(as.Date("2016/1/1"), as.Date("2020/12/31"), "days")
pd <- as_tibble(possible_dates) %>% 
  rename(date = value)

ans <- ord1 %>% 
  select(date) %>%
  distinct()

ans2 <- anti_join(pd, ans, by = "date")
ans2
```
Below I created a bar chart that shows the number of passengers that arrived at O'Hare for each month of each year:

```{r}
q3_1 <- ord1 %>%
  select(year, month, passengers) %>% 
  group_by(month, year) %>% 
  summarize(n = n(), passengers = sum(passengers)) %>% 
  select(year, month, passengers)
  
ggplot(data = q3_1) +
  geom_col(mapping = aes(x = month, y = passengers, fill = as.character(year)), position = "dodge") +
  scale_y_continuous(labels = label_comma()) +
  guides(fill = guide_legend(title="Year")) +
  xlab("Month") +
  ylab("Passengers") +
  ggtitle("O'Hare Airport Passengers")
```
We see that 2020 passenger totals were way lower than 2019 (due to COVID-19) and that there are the most passenger arrivals during the summer (Jun-Aug), peaking in July.

Then, I created a table to find the average number of passengers per day of the week for each year. I graphed the data first all together, then separately by year.

```{r}
q4 <- ord1 %>%
  select(wday, year, passengers) %>% 
  group_by(wday, year) %>% 
  summarize(n = n(), avg_passengers = sum(passengers) / n) %>% 
  select(-n)
q4  
 
graph <- ggplot(data = q4) +
  geom_col(mapping = aes(x = wday, y = avg_passengers, fill = as.character(year)), position = "dodge") +
  scale_y_continuous(labels = label_comma()) +
  guides(fill = guide_legend(title="Year")) +
  xlab("Day of the Week") +
  ylab("Average Passengers") +
  ggtitle("O'Hare Airport Passengers") +
  coord_flip() 
graph

graph2 <- graph + facet_wrap( ~ year, nrow = 2)
graph2
```
We can see that Saturday has the highest average number of passenger arrivals per year and Tuesday has the lowest.

In the following code, I:
- Added a season column
- Created a scatter plot of the average wait times of US and non-US passport holders
- Grouped the points by their season
- Added a constant slope line = 1 for each season
- Added regression line for each season

```{r}

q6 <- ord1 %>% 
  select(date, us_avg_wait, non_us_avg_wait) %>% 
  mutate(month = month(date)) %>%
  mutate(year = year(date)) %>% 
  mutate( Season = case_when(
      month == 3 | month == 4 | month == 5 ~ "Spring",
      month == 6 | month == 7 | month == 8 ~ "Summer",
      month == 9 | month == 10 | month == 11 ~ "Fall",
      month == 12 | month == 1 | month == 2 ~ "Winter")) %>% 
  select(-date, -month) 
q6

ggplot(data = q6, mapping = aes(x = us_avg_wait, y = non_us_avg_wait)) +
  xlab("Average US Passenger Wait Time") +
  ylab("Average Non-US Passenger Wait Time")+
  ggtitle("Non-US vs. US Passenger Wait Times") +
  geom_point(mapping = aes(color = Season), alpha = 1/2) + 
  geom_smooth(mapping = aes(color = Season),method = "lm") +
  geom_abline(mapping = aes(intercept = 0, slope = 1)) +
  facet_wrap( ~ year, nrow = 2)
```
We see the regression lines show a steady pattern that US passengers wait shorter times than non US citizens (every one of them has a slope > 1). Also, spring and summer typically have longer wait times than fall and winter.

The code below calculates the proportion of cases where the average time to get through customs is greater for non US passport holders than it is for passport holders:

```{r}
q7 <- ord1 %>% 
  select(date, us_avg_wait, non_us_avg_wait) %>% 
  mutate(year = year(date)) %>%
  filter(us_avg_wait < non_us_avg_wait) %>% 
  select(-date) %>% 
  group_by(year) %>% 
  summarize(n = n())

q7_total <- ord1 %>% 
  group_by(year) %>% 
  summarize(total = n())

ans <- full_join(q7, q7_total, by = "year") %>% 
  mutate(proportion = n/total) %>% 
  select(-n, -total)
ans
```
As seen above, proportion = cases where average wait time for non US citizens > average wait time of non-US citizens... These proportions contain values of y > x for each case in its respective year. It makes sense that on average folks who have a passport would get through customs faster than those without a passport.

Lastly, I:
- Added a column named booth_rate that estimates the average number of passengers per booth per hour
- Added a column named time_of_day which can be overnight, early morning, morning, afternoon, early evening, or late evening based on the hour variable
- Reordered the columns
- Filtered out cases with less than 200 passengers, a wait time of zero, or a booth rate above 500
- Created a boxplot of booth rates versus day of the week 
- Created another boxplot that displays the data by time of day as well

```{r}
q8 <- ord1 %>% 
  mutate(
    booth_rate = (passengers / booths) / (all_avg_wait/ 60))

q8_1 <- q8 %>% 
  mutate(
    time_of_day = case_when(
      hour == "0100 - 0200" | hour == "0200 - 0300" | hour == "0300 - 0400" | hour == "0400 - 0500" ~ "Overnight",
      hour == "0500 - 0600" | hour == "0600 - 0700" | hour == "0700 - 0800"  ~ "Early Morning",
      hour == "0800 - 0900" | hour == "0900 - 1000" | hour == "1000 - 1100" | hour == "1100 - 1200" ~ "Morning",
      hour == "1200 - 1300" | hour == "1300 - 1400" | hour == "1400 - 1500" | hour == "1500 - 1600" | hour == "1600 - 1700" ~ "Afternoon",
      hour == "1700 - 1800" | hour == "1800 - 1900" | hour == "1900 - 2000" ~ "Early Evening",
      hour == "2000 - 2100" | hour == "2100 - 2200" | hour == "2200 - 2300" | hour == "2300 - 0000" | hour == "0000 - 0100" ~ "Late Evening",)) %>% 
  filter(passengers > 199, all_avg_wait != 0, booth_rate < 500) %>%  
  select(wday, booth_rate, time_of_day) 
  
q8_1 <- q8_1 %>%
  mutate(i = case_when(
    time_of_day == "Overnight" ~ 1, 
    time_of_day == "Early Morning" ~ 2, 
    time_of_day == "Morning" ~ 3, 
    time_of_day == "Afternoon" ~ 4, 
    time_of_day == "Early Evening" ~ 5, 
    time_of_day == "Late Evening" ~ 6))

ans <- q8_1 %>% 
  arrange(i) %>% 
  select(wday, booth_rate, time_of_day)
ans

a <- ggplot(ans) +
  geom_boxplot(aes(x = wday, y = booth_rate, fill = wday)) +
  theme(axis.text.x = element_blank())+
  ylab("Booth Rate")+
  guides(fill = guide_legend(title="Day of Week"))
a
b <- a +
  facet_wrap( ~ time_of_day, nrow = 2) 
b
```
We see that Wednesday's have the slowest average booth rate for day of the week while early morning an average has the slowest booth rate for time of day.